Fast and scalable Gaussian process modeling with applications to astronomical time series \vspace{-3\baselineskip}

### 2017-01-12 paper/ms.tex

% Projects:
\newcommand{\project}[1]{\textsf{#1}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\genrp}{\project{genrp}}

@@

In the following pages, we will motivate the general problem of GP regression,
describe the previously published scalable method \citep{Rybicki:1995,
Ambikasaran:2015} and our generalization, and demonstrate the model's
application on various real and simulated data sets.
Alongside this paper, we have released efficient and well-tested
implementations of this method written in \project{C++}, \project{Python}, and
\project{Julia}.
These implementations are available online at \project{GitHub}
\url{https://github.com/dfm/GenRP} and \project{Zenodo} \dfmtodo{add zenodo
archive}.

%
% New Outline:
% 1. Introduction
% 2. Gaussian processes
%   - GPs in general
%   - RP method
%   - Ambikasaran method
%   - Generalization
%   - Relation to stochastically-driven, damped simple harmonic oscillator
% 3. Implementation and performance
%   - Implementation considerations: band solver vs. sparse, parameterization
%   - Ensuring positive-definite covariance matrix: Sturm's theorem
%   - Benchmark numbers
%   - Scaling with N and M
%   - Description of APIs
% 4. Examples
%   - Simulated data: a simple exact GenRP simulation, a GenRP simulation with
%     structure, a few standard kernels, a AGW kernel
%   - An AGN light curve
%   - An RR Lyrae
%   - A kepler light curve
%   - An asteroseismic target
% 5. Comparisons to other methods
% 6. Summary
%   - Possible applications
%

@@

In the following Section, we present a method of improving this scaling that
we call the \genrp\ method.
The \genrp\ method requires using a specific model for the covariance
$k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and it has several limitations.
The method can only be applied to one-dimensional datasets.
When we say ``one-dimensional'' here, it means that the \emph{input
coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv t_n$.\footnote{We are
using $t$ as the input coordinate because one-dimensional GPs are often
applied to time series data but this isn't a real restriction and the \genrp\
method can be applied to \emph{any} one-dimensional dataset.}
Furthermore, the covariance function for the \genrp\ method is ``stationary''.
This means that the function $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function
of $\tau_{nm} \equiv |t_n - t_m|$.

\section{The genrp model}

@@

It turns out that this kernel function can be made even more general by
introducing complex parameters $a_j \to a_j+i\,b_j$ and $c_j \to c_j+i\,d_j$.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{genrp-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be solved with
$\mathcal{O}(J^2\,N)$ operations.
The details of this method and a few implementation considerations are
discussed in \app{implementation}.

\eq{genrp-kernel-complex} can also be expressed as
\begin{eqnarray}\eqlabel{genrp-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
The Fourier transform of this covariance function is the power spectral
density (PSD) of the model and it is given by
\begin{eqnarray}\eqlabel{genrp-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
will return to a more general discussion of the physical intuition in a moment
but we can start with a discussion of some useful special cases of this model.

If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{genrp-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}\eqlabel{lorentz-psd}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
\right] \quad.
\end{eqnarray}
This PSD is the sum of two Lorentzian or Cauchy distributions with width
$c_j$ centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = d_j\,(2\,c_j)^{-1}$, and period
$P_j = 2\,\pi\,{d_j}^{-1}$.

@@

\section{Genrp as a model of stellar variations}

Another special case of the \genrp\ model of great physical interest is a
stochastically-driven simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
In the limit of an infinite time series and white random forcing, the PSD of
this equation is given by \citep{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
= \sqrt{2/\pi}\,S_0$.
The power spectrum in \eq{sho-psd} matches \eq{genrp-psd} if
\begin{eqnarray}
a_j &=& S_0\,\omega_0\,Q \\
    b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \genrp\
terms with parameters
\begin{eqnarray}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right] \\
d_{j\pm} &=& 0 \quad.
\end{eqnarray}

@@

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
    \sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function
\begin{eqnarray}\eqlabel{approx-matern}
k(\tau) &=& \lim_{f \to 0}\,
    S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
This suggests that the Mat\'ern-3/2 covariance can be well approximated using
the \genrp\ framework with a small value of $f$ in \eq{approx-matern} but we
caution that this could also lead to numerical issues with the solver.
}

@@

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we will take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \genrp\ method described
in the previous section.

@@

\begin{enumerate}
\item an OU process
\item a full GenRP process
\item Some standard kernels: Matern, Exp-squared, etc.
\item A KISS-GP process
\end{enumerate}

@@

\genrp\ alleviates these problems by providing a physically motivated
probabilistic model that can be evaluated efficiently even for large datasets.
In practice, we will model the star as a mixture of stochastically driven
simple harmonic oscillators where the amplitudes and frequencies of the
oscillations are computed using a physical model and evaluate the probability
of the observed dataset using a Gaussian Process with a PSD given by a sum of
terms given by \eq{sho-psd}.
This gives us a method of computing the likelihood function for the parameters
of the physical model in $\mathcal{O}(N)$ operations and this can be combined
with standard non-linear optimization or posterior sampling methods to infer
constraints on the parameters for the given dataset.

@@

Another source of variability that can be measured from time series
measurements of stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
(\dfmtodo{CITE}).
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star (\dfmtodo{CITE}) and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ in prep.).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the period of the oscillation.
The key difference between this function and other quasiperiodic kernels is
that it is everywhere positive.
We can construct a simple \genrp\ covariance function with similar properties
as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{a}{2+b}\,e^{-c\,\tau}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
\right]
\end{eqnarray}
for $a>0$, $b>0$, and $c>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a substantial gain in computational efficiency.

@@

Limitations of GenRP: one-dimension, stationary, etc.

@@

% A significant advance was made in generalizing the Press-Rybicki method to an
% arbitrary number of exponential kernels by utilizing the fact that the elements of
% such matrices can be written in terms of the products of components of two vectors,
% which are so-called semi-separable matrices \citep{Ambikasaran:2015}.  This `Generalized
% Rybicki-Press' (\texttt{GenRP}) approach
% has many advantages:  the computation time and storage both scale as $O(N_{time})$,
% the coefficients of the kernels need not be stationary, the computation of the
% likelihood is exact, and the algorithm is simple to implement, and thus is robust.
% Unlike the \texttt{CARMA} model, there is no restriction on the relation between
% the coefficients (except that the kernel needs to be positive definite).  Unlike
% \texttt{HODLR} and \texttt{KISS-GP},  the \texttt{GenRP} method is exact.  The main drawback
% of this approach is that the sum of exponential kernels with real coefficients can
% only represent a limited range of power spectra of auto-correlated time series:
% those which can be expressed the sum of Lorentzians that peak at zero frequency.

% However, this drawback turns out to not be significant:  the exponential kernel
% can also have a complex exponential coefficients (plus their complex-conjugates), and
% thus approximate an almost arbitrary autocorrelation function with enough terms, most
% importantly allowing for quasi-periodic kernels.  It turns out that the sum of Lorentzian
% components is a {\it very} good approximation of stellar variability spectra:  the
% activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
% Lorentzians \citep{Harvey:1985}, the asteroseismic oscillations may be expressed as non-zero frequency
% Lorentzians \citep{1990ApJ...364..699A,Gruberbauer:2009}, and stellar rotation may be
% parameterized this way as well.  In this paper
% we extend the \texttt{GenRP} approach to kernels with complex exponentials, which when
% combined with the complex conjugate leads to damped harmonic auto-correlation functions
% which are flexible enough to describe a wide range of variability, and may still
% be solved in $O(N_{time}$ operations.

@@

Although the complex kernel is a straightforward extension to the \texttt{GenRP} formalism, there is a disadvantage
in computational speed:  complex arithmetic requires $\approx 3$ times as many operations as real
arithmetic.  In addition, it turns out that the complex conjugate component, $(c_j+id_j)^* = c_j - id_j$, in
the extended matrix gives a solution which, not surprisingly, is the complex conjugate of the
component $c_j + i d_j$;  thus the computation is redundant.  Finally, complex
arithmetic has the disadvantage that it is not included in automatic-differentiation packages,
which is what we would like to use to compute the derivative of the likelihood (as it is very difficult
to compute this explicitly!).

@@

\vspace{1.5em}
All of the code used in this project is available from
\url{https://github.com/dfm/GenRP} under the MIT open-source software
license.
This code (plus some dependencies) can be run to re-generate all of the
figures and results in this paper; this version of the paper was generated
with git commit \texttt{\githash} (\gitdate).

@@

There are three primary applications we envision for the \genrp\ formalism: 1).\ modeling
the variability of an astrophysical system to infer it properties; 2).\
accounting for astrophysical variability as a source of noise when trying to detect
additional phenomena; 3).\ interpolating or extrapolating variability to future or missing
times.  Some examples of the first are determining the characteristic
variability timescale of a quasar, measuring the asteroseismic variation of a star,
or detection of quasi-periodic variability in a high-energy source.  Examples of the
second are detecting transiting exoplanets and correcting for stellar activity in
radial velocity measurements. Examples of the third are measuring time delays in
multiply-imaged gravitationally lensed sources and reverberation mapping of AGN.

@@

Given the drawbacks of these approaches, the \genrp\ formalism allows both a fast
computation of the likelihood in ${\cal O}(N)$ time, as well as a functional form
that accurately describes stellar variability due to the relation to the simple
harmonic oscillator power spectrum, which is an analog of stellar asteroseismic
oscillations.  As higher signal-to-noise observations of transiting exoplanet systems
are obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters, and so we expect that \genrp\
will be important for transit timing, transit spectroscopy, Doppler beaming,
phase functions, and more.

@@

\bibliography{genrp}

>>>>

% Projects:
\newcommand{\project}[1]{\textsf{#1}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\celerite}{\project{celerite}}

@@

In the following pages, we will motivate the general problem of GP regression,
describe the previously published scalable method \citep{Rybicki:1995,
Ambikasaran:2015} and our generalization, and demonstrate the model's
application on various real and simulated data sets.
Alongside this paper, we have released efficient and well-tested
implementations of this method written in \project{C++}, \project{Python}, and
\project{Julia}.
These implementations are available online at \project{GitHub}
\url{https://github.com/dfm/celerite} and \project{Zenodo} \dfmtodo{add zenodo
archive}.

%
% New Outline:
% 1. Introduction
% 2. Gaussian processes
%   - GPs in general
%   - RP method
%   - Ambikasaran method
%   - Generalization
%   - Relation to stochastically-driven, damped simple harmonic oscillator
% 3. Implementation and performance
%   - Implementation considerations: band solver vs. sparse, parameterization
%   - Ensuring positive-definite covariance matrix: Sturm's theorem
%   - Benchmark numbers
%   - Scaling with N and M
%   - Description of APIs
% 4. Examples
%   - Simulated data: a simple exact celerite simulation, a celerite simulation with
%     structure, a few standard kernels, a AGW kernel
%   - An AGN light curve
%   - An RR Lyrae
%   - A kepler light curve
%   - An asteroseismic target
% 5. Comparisons to other methods
% 6. Summary
%   - Possible applications
%

@@

In the following Section, we present a method of improving this scaling that
we call the \celerite\ method.
The \celerite\ method requires using a specific model for the covariance
$k_\bvec{\alpha}(\bvec{x}_n,\,\bvec{x}_m)$ and it has several limitations.
The method can only be applied to one-dimensional datasets.
When we say ``one-dimensional'' here, it means that the \emph{input
coordinates} $\bvec{x}_n$ are scalar, $\bvec{x}_n \equiv t_n$.\footnote{We are
using $t$ as the input coordinate because one-dimensional GPs are often
applied to time series data but this isn't a real restriction and the \celerite\
method can be applied to \emph{any} one-dimensional dataset.}
Furthermore, the covariance function for the \celerite\ method is ``stationary''.
This means that the function $k_\bvec{\alpha}(t_n,\,t_m)$ is only a function
of $\tau_{nm} \equiv |t_n - t_m|$.

\section{The celerite model}

@@

It turns out that this kernel function can be made even more general by
introducing complex parameters $a_j \to a_j+i\,b_j$ and $c_j \to c_j+i\,d_j$.
In this case, the covariance function becomes
\begin{eqnarray}\eqlabel{celerite-kernel-complex}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    \frac{1}{2}(a_j + i\,b_j)\,\exp\left(-(c_j+i\,d_j)\,\tau_{nm}\right)
        \right. \nonumber\\
    &&+\left.
    \frac{1}{2}(a_j - i\,b_j)\,\exp\left(-(c_j-i\,d_j)\,\tau_{nm}\right)
\right]
\end{eqnarray}
and, for this function, \eq{gp-likelihood} can still be solved with
$\mathcal{O}(J^2\,N)$ operations.
The details of this method and a few implementation considerations are
discussed in \app{implementation}.

\eq{celerite-kernel-complex} can also be expressed as
\begin{eqnarray}\eqlabel{celerite-kernel}
k_\bvec{\alpha}(\tau_{nm}) = \sigma_n^2\,\delta_{nm} +
    \sum_{j=1}^J &&\left[
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
        \right.\nonumber\\
    &&+ \left.
    b_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\sin\left(d_j\,\tau_{nm}\right)
\right] \quad.
\end{eqnarray}
The Fourier transform of this covariance function is the power spectral
density (PSD) of the model and it is given by
\begin{eqnarray}\eqlabel{celerite-psd}
S(\omega) = \sum_{j=1}^J \sqrt{\frac{2}{\pi}}
\frac{(a_j\,c_j+b_j\,d_j)\,({c_j}^2+{d_j}^2)+(a_j\,c_j-b_j\,d_j)\,\omega^2}
{\omega^4+2\,({c_j}^2-{d_j}^2)\,\omega^2+({c_j}^2+{d_j}^2)^2}\quad.
\end{eqnarray}
The physical interpretation of this model isn't immediately obvious and we
will return to a more general discussion of the physical intuition in a moment
but we can start with a discussion of some useful special cases of this model.

If we set the imaginary amplitude $b_j$ for some component $j$ to zero, that
term of \eq{celerite-kernel} becomes
\begin{eqnarray}
k_j(\tau_{nm}) =
    a_j\,\exp\left(-c_j\,\tau_{nm}\right)\,\cos\left(d_j\,\tau_{nm}\right)
\end{eqnarray}
and the PSD for the this component is
\begin{eqnarray}\eqlabel{lorentz-psd}
S_j(\omega) = \frac{1}{\sqrt{2\,\pi}}\,\frac{a_j}{c_j}\,\left[
    \frac{1}{1+\left(\frac{\omega-d_j}{c_j}\right)^2} +
    \frac{1}{1+\left(\frac{\omega+d_j}{c_j}\right)^2}
\right] \quad.
\end{eqnarray}
This PSD is the sum of two Lorentzian or Cauchy distributions with width
$c_j$ centered on $\omega = \pm d_j$.
This model can be interpreted intuitively as a quasiperiodic oscillator with
amplitude $A_j = a_j$, quality factor $Q_j = d_j\,(2\,c_j)^{-1}$, and period
$P_j = 2\,\pi\,{d_j}^{-1}$.

@@

\section{celerite as a model of stellar variations}

Another special case of the \celerite\ model of great physical interest is a
stochastically-driven simple harmonic oscillator.
The differential equation for this system is
\begin{equation}
    \left[\frac{\dd^2}{\dd t^2} + \frac{\omega_0}{Q}\,\frac{\dd}{\dd t}
    + \omega_0^2\right]\, y(t) = \epsilon(t)
\end{equation}
where $\omega_0$ is the frequency of the undamped oscillator, $Q$ is the
quality factor of the oscillator, and $\epsilon(t)$ is a stochastic driving
force.
In the limit of an infinite time series and white random forcing, the PSD of
this equation is given by \citep{Anderson:1990}
\begin{equation}\eqlabel{sho-psd}
S(\omega) = \sqrt{\frac{2}{\pi}}\,\frac{S_0\,\omega_0^4}
    {(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2}
\end{equation}
where $S_0$ is proportional to the power at $\omega = \omega_0$, $S(\omega_0)
= \sqrt{2/\pi}\,S_0$.
The power spectrum in \eq{sho-psd} matches \eq{celerite-psd} if
\begin{eqnarray}
a_j &=& S_0\,\omega_0\,Q \\
    b_j &=& \frac{S_0\,\omega_0\,Q}{\sqrt{4\,Q^2-1}} \\
c_j &=& \frac{\omega_0}{2\,Q}\\
d_j &=& \frac{\omega_0}{2\,Q} \sqrt{4\,Q^2-1} \quad,
\end{eqnarray}
for $Q \ge \frac{1}{2}$.
For $0 < Q \le \frac{1}{2}$, \eq{sho-psd} can be captured by a pair of \celerite\
terms with parameters
\begin{eqnarray}
a_{j\pm} &=& \frac{1}{2}\,S_0\,\omega_0\,Q\,\left[ 1 \pm
        \frac{1}{\sqrt{1-4\,Q^2}}\right] \\
b_{j\pm} &=& 0 \\
    c_{j\pm} &=& \frac{\omega_0}{2\,Q}\,\left[1 \mp \sqrt{1-4\,Q^2}\right] \\
d_{j\pm} &=& 0 \quad.
\end{eqnarray}

@@

{\item Substituting $Q = 1/2$, \eq{sho-psd} becomes
\begin{eqnarray}
S(\omega) =
    \sqrt{\frac{2}{\pi}}\,\frac{S_0}{\left[(\omega/\omega_0)^2+1\right]^2}
\end{eqnarray}
with the corresponding covariance function
\begin{eqnarray}\eqlabel{approx-matern}
k(\tau) &=& \lim_{f \to 0}\,
    S_0\,\omega_0\,e^{-\omega_0\,\tau}\,
    \left[\cos(f\,\tau) + \frac{\omega_0}{f}\,\sin(f\,\tau)\right] \\
&=& S_0\,\omega_0\,e^{-\omega_0\,\tau}\,[1+\omega_0\,\tau] \quad.
\end{eqnarray}
This covariance function is also known as the Mat\'ern-3/2 function
\citep{Rasmussen:2006}.
This suggests that the Mat\'ern-3/2 covariance can be well approximated using
the \celerite\ framework with a small value of $f$ in \eq{approx-matern} but we
caution that this could also lead to numerical issues with the solver.
}

@@

These special cases demonstrate that the stochastically-driven simple harmonic
oscillator provides a physically motivated model that is flexible enough to
describe a wide range of stellar variations.
Low $Q \approx 1$ can capture granulation noise and high $Q \gg 1$ is a good
model for asteroseismic oscillations.
In practice, we will take a sum over oscillators with different values of $Q$,
$S_0$, and $\omega_0$ to give a sufficient accounting of the power spectrum
stellar time series.
Since this kernel is exactly described by the exponential kernel, the
likelihood (\eqalt{gp-likelihood}) can be evaluated for a time series with $N$
measurements in $\mathcal{O}(N)$ operations using the \celerite\ method described
in the previous section.

@@

\begin{enumerate}
\item an OU process
\item a full celerite process
\item Some standard kernels: Matern, Exp-squared, etc.
\item A KISS-GP process
\end{enumerate}

@@

\celerite\ alleviates these problems by providing a physically motivated
probabilistic model that can be evaluated efficiently even for large datasets.
In practice, we will model the star as a mixture of stochastically driven
simple harmonic oscillators where the amplitudes and frequencies of the
oscillations are computed using a physical model and evaluate the probability
of the observed dataset using a Gaussian Process with a PSD given by a sum of
terms given by \eq{sho-psd}.
This gives us a method of computing the likelihood function for the parameters
of the physical model in $\mathcal{O}(N)$ operations and this can be combined
with standard non-linear optimization or posterior sampling methods to infer
constraints on the parameters for the given dataset.

@@

Another source of variability that can be measured from time series
measurements of stars is rotation.
The inhomogeneous surface of the star (spots, plage, \etc) imprints itself as
quasiperiodic variations in photometric or spectroscopic observations
(\dfmtodo{CITE}).
It has been demonstrated that for light curves with nearly uniform sampling,
the empirical autocorrelation function provides a reliable estimate of the
rotation period of a star (\dfmtodo{CITE}) and that a GP model with a
quasiperiodic covariance function can be used to make probabilistic
measurements even with sparsely sampled data (R.~Angus, \etal\ in prep.).
The covariance function used for this type of analysis has the form
\begin{eqnarray}\eqlabel{sine2}
k(\tau) = A\,\exp\left(-\frac{\tau^2}{2\,\ell^2} -
    \Gamma\,\sin^2\left(\frac{\pi\,\tau}{P} \right) \right)
\end{eqnarray}
where $P$ is the period of the oscillation.
The key difference between this function and other quasiperiodic kernels is
that it is everywhere positive.
We can construct a simple \celerite\ covariance function with similar properties
as follows
\begin{eqnarray}\eqlabel{rot-kernel}
k(\tau) = \frac{a}{2+b}\,e^{-c\,\tau}\,\left[
    \cos\left(\frac{2\,\pi\,\tau}{P}\right) + (1 + b)
\right]
\end{eqnarray}
for $a>0$, $b>0$, and $c>0$.
The covariance function in \eq{rot-kernel} cannot exactly reproduce \eq{sine2}
but, since \eq{sine2} is only an effective model, \eq{rot-kernel} can be used
as a drop-in replacement for a substantial gain in computational efficiency.

@@

Limitations of celerite: one-dimension, stationary, etc.

@@

% A significant advance was made in generalizing the Press-Rybicki method to an
% arbitrary number of exponential kernels by utilizing the fact that the elements of
% such matrices can be written in terms of the products of components of two vectors,
% which are so-called semi-separable matrices \citep{Ambikasaran:2015}.  This `Generalized
% Rybicki-Press' (\texttt{celerite}) approach
% has many advantages:  the computation time and storage both scale as $O(N_{time})$,
% the coefficients of the kernels need not be stationary, the computation of the
% likelihood is exact, and the algorithm is simple to implement, and thus is robust.
% Unlike the \texttt{CARMA} model, there is no restriction on the relation between
% the coefficients (except that the kernel needs to be positive definite).  Unlike
% \texttt{HODLR} and \texttt{KISS-GP},  the \texttt{celerite} method is exact.  The main drawback
% of this approach is that the sum of exponential kernels with real coefficients can
% only represent a limited range of power spectra of auto-correlated time series:
% those which can be expressed the sum of Lorentzians that peak at zero frequency.

% However, this drawback turns out to not be significant:  the exponential kernel
% can also have a complex exponential coefficients (plus their complex-conjugates), and
% thus approximate an almost arbitrary autocorrelation function with enough terms, most
% importantly allowing for quasi-periodic kernels.  It turns out that the sum of Lorentzian
% components is a {\it very} good approximation of stellar variability spectra:  the
% activity and granulation noise due to stellar convection may be expressed in terms of zero-frequency
% Lorentzians \citep{Harvey:1985}, the asteroseismic oscillations may be expressed as non-zero frequency
% Lorentzians \citep{1990ApJ...364..699A,Gruberbauer:2009}, and stellar rotation may be
% parameterized this way as well.  In this paper
% we extend the \texttt{celerite} approach to kernels with complex exponentials, which when
% combined with the complex conjugate leads to damped harmonic auto-correlation functions
% which are flexible enough to describe a wide range of variability, and may still
% be solved in $O(N_{time}$ operations.

@@

Although the complex kernel is a straightforward extension to the \texttt{celerite} formalism, there is a disadvantage
in computational speed:  complex arithmetic requires $\approx 3$ times as many operations as real
arithmetic.  In addition, it turns out that the complex conjugate component, $(c_j+id_j)^* = c_j - id_j$, in
the extended matrix gives a solution which, not surprisingly, is the complex conjugate of the
component $c_j + i d_j$;  thus the computation is redundant.  Finally, complex
arithmetic has the disadvantage that it is not included in automatic-differentiation packages,
which is what we would like to use to compute the derivative of the likelihood (as it is very difficult
to compute this explicitly!).

@@

\vspace{1.5em}
All of the code used in this project is available from
\url{https://github.com/dfm/celerite} under the MIT open-source software
license.
This code (plus some dependencies) can be run to re-generate all of the
figures and results in this paper; this version of the paper was generated
with git commit \texttt{\githash} (\gitdate).

@@

There are three primary applications we envision for the \celerite\ formalism: 1).\ modeling
the variability of an astrophysical system to infer it properties; 2).\
accounting for astrophysical variability as a source of noise when trying to detect
additional phenomena; 3).\ interpolating or extrapolating variability to future or missing
times.  Some examples of the first are determining the characteristic
variability timescale of a quasar, measuring the asteroseismic variation of a star,
or detection of quasi-periodic variability in a high-energy source.  Examples of the
second are detecting transiting exoplanets and correcting for stellar activity in
radial velocity measurements. Examples of the third are measuring time delays in
multiply-imaged gravitationally lensed sources and reverberation mapping of AGN.

@@

Given the drawbacks of these approaches, the \celerite\ formalism allows both a fast
computation of the likelihood in ${\cal O}(N)$ time, as well as a functional form
that accurately describes stellar variability due to the relation to the simple
harmonic oscillator power spectrum, which is an analog of stellar asteroseismic
oscillations.  As higher signal-to-noise observations of transiting exoplanet systems
are obtained, the effects of stellar variability will more dramatically impact the
correct inference of planetary transit parameters, and so we expect that \celerite\
will be important for transit timing, transit spectroscopy, Doppler beaming,
phase functions, and more.

@@

\bibliography{celerite}
